---
title: "NYC Shooting Project"
author: "student"
date: "7/12/2021"
output:
  pdf_document: default
---



### NYC Shooting:

This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. (source: <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>)


### Questions:

1) What is the trend with regard to years?  
2) What is the lowest/highest number of incidents per victim age group? 

### Bias:
Coming from New York, by bias was based on pride. Finding out what borough had the highest number was my focus and the age groups within that area. I think I wanted to prove the area that I grew up in had a lower incident rate than others. In regards to selecting the data, well, that was the assignment, but I asked the question if I would do this on my own, and yes, I would – so that needs to be accounted for as well. Even though this RMD file does not have further studies, I reviewed later, other factors, including race, age, and murder count. These to have a question of bias attached to them.

Note: Other bias reasons: A byproduct of limited knowledge of the field of analytics & knowledge of R, which is the reason for taking said class.



### Loading R Packages:
```{r library, echo=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2) # - part of tidyverse
library(dplyr) # - part of tidyverse
```


### Reading in the Data:
```{r reading, echo=TRUE}
#url_in <-"https://data.cityofnewyork.us/resource/833y-fsy8.csv" - appears to be an issue, seek other path
url_in <-"https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
dataShooting <- read.csv(url_in)
```

### Summary of Data:
```{r reviwing, echo=TRUE}
summary(dataShooting)
head(dataShooting)
# DETAILS OF SUMMARY:
  # INCIDENT_KEY: provides a unique key/id to every row
  # OCCUR_DATE: provides date of incident
  # OCCUR_TIME: provides time of incident
  # BORO: provides the location per boro (five boros to NY City)
  # PRECINCT: provides the police station reasonable 
  # JURISDICTION_CODE: provides police station location and j code
  # LOCATION_DESC: provides location of incident
  # STATISTICAL_MURDER_FLAG: provides details of shooting (fatal)
  # PERP_AGE_GROUP, PERP_SEX, PERP_RACE: provides details of perpetrator
  # VIC_AGE_GROUP, VIC_SEX, VIC_RACE: provides details of victim
  # X_COORD_CD, Y_COORD_CD, LATITUDE, LONGITUDE, GEOCODED_COLUMN: provides location information of incident

```

### # TIDY DATA (a wee bit):
```{r cleaning, echo=TRUE}

#dataShooting <- separate(dataShooting,occur_date, into = c("date", "timeRemove"), sep = "T") - basic, works, but issues with testing
#dataShooting$BORO <- as.factor(dataShooting$BORO) - factor works, but issues with plots, dropped!

dataShooting <- mutate(dataShooting, OCCUR_DATE = mdy(OCCUR_DATE), dsSHOOTINGS = 1,
                       YEARS = format_ISO8601(OCCUR_DATE, precision = "y"),
                       YEAR_MONTHS = format_ISO8601(OCCUR_DATE, precision = "ym"))

```
### SELECTING COLUMNS:
```{r selectingcolumns, echo=TRUE}
dsPlotReview <- select(dataShooting, OCCUR_DATE, VIC_AGE_GROUP, YEAR_MONTHS, YEARS, OCCUR_TIME, BORO, dsSHOOTINGS) %>%
  filter(dsSHOOTINGS == "1") 
```

### Three Basic Visualizations:
```{r visual, echo=TRUE}

dsPlotOne <- dsPlotReview %>%
  group_by(BORO) %>%
  count()

dsPlotViewOne <-ggplot(dsPlotOne, aes(x = BORO, weight =n)) +
  geom_bar(width = .75, fill = "blue") +
  labs(x = "City of NYC: Boroughs", y = "Number of Shootings", 
       title = "A View by NYC Boroughs")
plot(dsPlotViewOne)

#--------------------------------------------------------------
dsPlotTwo <- dsPlotReview %>%
  group_by(YEARS) %>%
  count()

dsPlotViewTwo <-ggplot(dsPlotTwo, aes(x = YEARS, weight =n)) +
  geom_bar(width = .75, fill = "purple") +
  labs(x = "Years: 2006 to 2020", y = "Number of Shootings", 
       title = "A View by The Years")
plot(dsPlotViewTwo)

#--------------------------------------------------------------
dsPlotThree <- dsPlotReview %>%
  group_by(VIC_AGE_GROUP) %>%
  filter(VIC_AGE_GROUP != "UNKNOWN") %>%
  count()

dsPlotViewThree <-ggplot(dsPlotThree, aes(x = VIC_AGE_GROUP, weight =n)) +
  geom_bar(width = .75, fill = "orange") +
  labs(x = "Victims Age Group", y = "Number of Shootings", 
       title = "A View by Victims Age Group")
plot(dsPlotViewThree)

```

## Conclusion:

### Review of Bias:
Coming from New York, by bias was based on pride. Finding out what borough had the highest number was my focus and the age groups within that area. I think I wanted to prove the area that I grew up in had a lower incident rate than others. In regards to selecting the data, well, that was the assignment, but I asked the question if I would do this on my own, and yes, I would – so that needs to be accounted for as well. Even though this RMD file does not have further studies, I reviewed later, other factors, including race, age, and murder count. These to have a question of bias attached to them.

Note: Other bias reasons: A byproduct of limited knowledge of the field of analytics & knowledge of R, which is the reason for taking said class.

### Question & Answers: 
1) What is the trend with regard to years? - According to the data, incidents were going down, but made a sharp turn up during the past year (2020).  

2) What is the lowest/highest number of incidents per victim age group?  - According to the data, the age group of "25-44" has the greatest number incidents. 

